### 2015-05-22 Harris_2015/manuscript-materials/draft.tex
actually resubmitted

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in our data sets (MacArthur 1958, Diamond 1975). Despite
decades of work and several major controversies, however (Lewin 1983,
Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009), and thus important
ecological processes remain poorly understood. More generally, it can be
difficult to reason about the complex web of direct and indirect species
interactions in real assemblages, especially when these interactions
occur against a background of other ecological processes such as
dispersal and environmental filtering (Connor et al. 2013). For this
reason, it isn't always clear what kinds of patterns would even
constitute evidence of competition, as opposed to some other biological
process or random sampling error (Lewin 1983, Roughgarden 1983).

@@

While the competition between the two shrubs in the previous example
does not leave the commonly-expected pattern in community structure
(negative association at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, \emph{among shaded
sites}, there will be a deficit of co-occurrences, and \emph{among
unshaded sites}, there will also be such a deficit.

@@

The rest of the paper proceeds as follows. First, I discuss how Markov
networks work and how they can be used to simulate landscape-level data
or to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

@@

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
ensures that the probabilities of all possible assemblages that could be
produced by the model sum to one (bottom of Figure 2B). Calculating
\(Z(\alpha, \beta)\) exactly, as is done in this paper, quickly becomes
infeasible as the number of species increases: with \(2^N\) possible
assemblages of \(N\) species, the number of bookkeeping operations
required for exact inference spirals exponentially into the billions and
beyond. Numerous techniques are available for working with Markov
networks that keep the computations tractable, e.g.~via analytic
approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but they are beyond the scope of this paper.

@@

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} The previous sections involved known values of
\(\alpha\) and \(\beta\). In practice, ecologists will often need to
estimate these parameters from data instead. When the number of species
is reasonably small, one can compute exact maximum likelihood estimates
for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum is the unique combination of \(\alpha\) and \(\beta\)
coefficients that would be expected to produce exactly the observed
co-occurrence frequencies. For the analyses in this paper, I used the
\emph{rosalia} package (Harris 2015a) for the R programming language (R
Core Team 2015) to define the objective function and gradient as R code.
The rosalia package then uses the \texttt{BFGS} method in R's
\texttt{optim} function to find the best values for \(\alpha\) and
\(\beta\).

@@

The first two interaction measures were the sample covariances and the
partial covariances between each pair of species' data vectors on the
landscape (Albrecht and Gotelli 2001). Because partial covariances are
undefined for landscapes with perfectly-correlated species pairs, I used
a regularized estimate based on ridge regression {[}Wieringen and
Peeters (2014); i.e.~linear regression with a Gaussian prior{]}. For
these analyses, I set the ridge parameter to 0.2 divided by the number
of sites on the landscape.

The third method, described in Gotelli and Ulrich (2009), involved
simulating possible landscapes from a null model that retains the row
and column sums of the original matrix (Strong et al. 1984). Using the
default options in the Pairs software described in Gotelli and Ulrich
(2009), I simulated the null distribution of scaled C-scores (a test
statistic describing the number of \emph{non}-co-occurrences between two
species). The software then calculated a \(Z\) statistic for each
species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

\textbf{\emph{Method evaluation.}} I evaluated each method qualitatively
on the simulated landscapes based on Figure 1: any method that reliably
determined that the two shrub species were negatively associated passed;
other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate. For each method, I calculated the squared deviations
between each species pair's rescaled estimate and the ``true''
interactions used to generate the assemblages. Finally, I averaged these
squared errors for each combination of species richness, landscape size,
statistical method, and presence/absence of environmental filtering
across all 12 replicates; the mean squared errors associated with these
subsets of the data determined the proportion of variance explained by
each method under different conditions.

@@

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level--- due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations on 94\% of simulated landscapes, and the
randomization-based null model falsely reported such associations 100\%
of the time. The two methods for evaluating conditional relationships
(Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each
correctly identified the direct negative interaction between the shrubs
94\% of the time.

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: overall, the Markov network explained 54\% of the
variance in the ``true'' parameters, followed by partial covariances
(33\%), and sample covariances (22\%).

The null model scores initially explained only 12\% of the variation.
After manually reducing the value of one especially strong outlier
(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%
(Appendix B). Figure 3 reflects the adjusted version of the results.

@@

The results presented above show that Markov networks, unlike a common
null modeling approach, can recover species' pairwise interactions from
observational data. The Markov networks were successful even when direct
interactions are largely overwhelmed by indirect effects (Figure 1) or
environmental effects (lower panels of Figure 3). For cases where
fitting a Markov network is computationally infeasible, these results
also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide a
surprisingly useful approximation. The success of the partial
correlation's success on simulated data may not carry over to real data
sets, however; Loh and Wainwright (2013) suggest that the linear
approximations may be less reliable in cases where the true interaction
matrix contains more structure (e.g.~guilds or trophic levels). On the
other hand, if ecologists are familiar enough with the natural history
of their study systems to describe this kind of structure as a prior
distribution on the parameters or as a penalty on the likelihood, then
real-world results might be even better than those shown in Figure 3.

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about
species interactions. The small simulated data sets described by Figure
1 show that test statistics based on marginal co-occurrence (such as
C-scores) will not always have a straightforward relationship with the
underlying ecological processes. More generally, deviations from the
null model generally provided less information about direct species
interactions than correlation coefficients did. Scientists currently
using null modeling approaches in their research may be able to capture
twice as much of the variance in species' true interaction strengths
using partial covariances from linear regression instead, or three times
as much using a Markov network.

Null and neutral models can be very useful for clarifying our thinking
about the numerical consequences of species' richness and abundance
patterns (Harris et al. 2011, Xiao et al. 2015), but deviations from a
null model must be interpreted with care (Roughgarden 1983). In complex
networks of ecological interactions---and even in small networks with
three species---it may simply not be possible to implicate individual
species pairs or specific ecological processes like competition by
rejecting a general- purpose null. Direct estimates of species'
conditional associations may be the only way to make these inferences
reliably.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in our data sets (MacArthur 1958, Diamond 1975). Despite
decades of work and several major controversies, however (Lewin 1983,
Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009), and thus important
ecological processes remain poorly understood. More generally, it can be
difficult to reason about the complex web of direct and indirect species
interactions in real assemblages, especially when these interactions
occur against a background of other ecological processes such as
dispersal limitation and environmental filtering (Connor et al. 2013).
For this reason, it isn't always clear what kinds of patterns would even
constitute evidence of competition, as opposed to some other biological
process or random sampling error (Lewin 1983, Roughgarden 1983).

@@

While competition between the two shrubs in the previous example does
not leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal
in the data (Figure 1C). Specifically, \emph{among shaded sites}, there
will be a deficit of co-occurrences, and \emph{among unshaded sites},
there will also be such a deficit.

@@

The rest of the paper proceeds as follows. First, I discuss how Markov
networks operate and how they can be used to simulate landscape-level
data or to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

@@

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which is
defined so that that the probabilities of all possible assemblages that
could be produced by the model sum to one (bottom of Figure 2B).
Calculating \(Z(\alpha, \beta)\) exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
\(2^N\) possible assemblages of \(N\) species, the number of bookkeeping
operations required for exact inference spirals exponentially into the
billions and beyond. Numerous techniques are available for working with
Markov networks that keep the computations tractable, e.g.~via analytic
approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but they are beyond the scope of this paper.

@@

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} In the previous two sections, the values of
\(\alpha\) and \(\beta\) were known. In practice, however, ecologists
will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact
maximum likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients by optimizing \(p(\vec{y}; \alpha, \beta)\). Fully-observed
Markov networks like the ones considered here have unimodal likelihood
surfaces (Murphy 2012), ensuring that this procedure will always
converge on the global maximum. This maximum is the unique combination
of \(\alpha\) and \(\beta\) coefficients that would be expected to
produce exactly the observed co-occurrence frequencies. For the analyses
in this paper, I used the \emph{rosalia} package (Harris 2015a) for the
R programming language (R Core Team 2015) to define the objective
function and gradient as R code. The rosalia package then uses the
\texttt{BFGS} method in R's \texttt{optim} function to find the best
values for \(\alpha\) and \(\beta\).

@@

The first two alternative interaction measures were the sample
covariances and the partial covariances between each pair of species'
data vectors on the landscape (Albrecht and Gotelli 2001). Because
partial covariances are undefined for landscapes with
perfectly-correlated species pairs, I used a regularized estimate based
on ridge regression {[}Wieringen and Peeters (2014); i.e.~linear
regression with a Gaussian prior{]}. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third alternative method, described in Gotelli and Ulrich (2009),
involved simulating possible landscapes from a null model that retains
the row and column sums of the original matrix (Strong et al. 1984).
Using the default options in the Pairs software described in Gotelli and
Ulrich (2009), I simulated the null distribution of scaled C-scores (a
test statistic describing the number of \emph{non}-co-occurrences
between two species). The software then calculated a \(Z\) statistic for
each species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

\textbf{\emph{Method evaluation.}} For the simulated landscapes based on
Figure 1, method evaluation was fairly qualitative: any method whose
test statistic for the two shrubs indicated a negative relationship
passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each method, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate. These regressions yielded squared errors for each of
the 23,140 simulated species pairs, across all conditions and
replicates. I then partitioned these errors into groups defined by the
properties of the simulated landscapes (i.e.~species richness, number of
observed communities, and the presence/absence of environmental
filtering). Within each partition, the mean squared error was used to
calculate the proportion of variance explained (compared with a baseline
model that assumed all interaction strengths to be zero).

@@

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level---due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations for 94\% of the simulated landscapes, and
the randomization-based null model falsely reported such associations
100\% of the time. The two methods for evaluating conditional
relationships (Markov networks and partial covariances), however,
successfully controlled for the indirect pathway via the tree species
and each correctly identified the direct negative interaction between
the shrubs 94\% of the time.

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: overall, the Markov network explained 54\% of the
``true'' parameters' squared deviations from zero, followed by partial
covariances (33\%), and sample covariances (22\%). The null model scores
initially explained only 12\%. After manually reducing the value of one
especially strong outlier (\(Z=1004\), implying \(p<10^{-1000000}\)),
this increased to 17\% (Appendix B). Figure 3 reflects the adjusted
version of the results.

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where a common null modeling technique reliably fails.
Specifically, Markov networks were successful even when direct
interactions were largely overwhelmed by indirect effects (Figure 1) or
environmental effects (lower panels of Figure 3). For cases where
fitting a Markov network is computationally infeasible, these results
also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide a
surprisingly useful approximation. The partial correlations' success on
simulated data may not carry over to real data sets, however; Loh and
Wainwright (2013) show that the linear approximations can be less
reliable in cases where the true interaction matrix contains more
structure (e.g.~guilds or trophic levels). On the other hand, if
ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on
the parameters or as a penalty on the likelihood, then this information
could reduce the effective degres of freedom to estimate and real-world
results might be even better than those shown in Figure 3.

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about
species interactions. Null and neutral models can be very useful for
clarifying our thinking about the numerical consequences of species'
richness and abundance patterns (Harris et al. 2011, Xiao et al. 2015),
but deviations from a null model must be interpreted with care
(Roughgarden 1983). In complex networks of ecological interactions (and
even in small networks with three species), it may simply not be
possible to implicate individual species pairs or specific ecological
processes like competition by rejecting a general-purpose null (Gotelli
and Ulrich 2009). Estimating pairwise coefficients directly seems like a
much more promising approach: to the extent that the models' relative
performance on real data sets is similar to the range of results shown
in Figure 3, scientists in this field could easily double their
explanatory power by switching from null models to linear regression and
partial covariances, or triple it by switching to a Markov network.
